---
title: "R Markdown GG Plot Week 5"
author: "Jennifer Kidson"
date: "February 5, 2016"
output: html_document
---

## joining data

For this portion of the individual assignment, similar to last week, you'll find it easiest to copy and paste from `## joining data` onwards in  [`env-info/wk05_joining.Rmd`](https://raw.githubusercontent.com/ucsb-bren/env-info/gh-pages/wk05_joining.Rmd) to your `env-info_hw/wk05_ggplot.Rmd`. Then you can play with different chunks of the code. Please be sure to answer all tasks and questions at the bottom.

The R chunks explaining the `dplyr` join functions below are pulled from the excellent [**wrangling-webinar.pdf**](wk03_dplyr/wrangling-webinar.pdf) presentation, which you should consult as you execute (see shortcuts in [rstudio-IDE-cheatsheet.pdf](refs/cheatsheets/rstudio-IDE-cheatsheet.pdf)). 

### setup

Ensure that you're in the same working directory `env-info_hw` when you Knit HTML as when you test code in the Console.

```{r setwd students}
wd = 'env-info_hw'

# set working directory for Console (vs Rmd)
if (wd %in% list.files() & interactive()){
    setwd(wd)
}

# ensure working directory
if (basename(getwd()) != wd){
  stop(sprintf("WHOAH! Your working directory is not in '%s'!\n   getwd(): %s", wd, getwd()))
}
```

### `bind_cols` 
(Just line up b next to a)

```{r bind_cols}

library(dplyr)
y = data.frame(
  x1 = c('A','B','C'), 
  x2 = c( 1 , 2 , 3), 
  stringsAsFactors=F)
z = data.frame(
  x1 = c('B','C','D'), 
  x2 = c( 2 , 3 , 4), 
  stringsAsFactors=F)
y
z
bind_cols(y, z)
```

### `bind_rows` 
(just stack a on top of b)

```{r bind_rows}
y
z
bind_rows(y, z)
```

### `union` 
(a+b, without duplicates)

```{r union}
y
z
union(y, z)
```

### `intersect` 
(common lines between a and b)

```{r intersect}
y
z
intersect(y, z)
```

### `setdiff` 
(lines of a and b that don't appear in the other column)

```{r setdiff}
y
z
setdiff(y, z)
```

### `left_join` 
(retain all the songs, and join to artists - drop artists that don't appear in songs)

```{r left_join}
songs = data.frame(
  song = c('Across the Universe','Come Together', 'Hello, Goodbye', 'Peggy Sue'),
  name = c('John','John','Paul','Buddy'), 
  stringsAsFactors=F)
artists = data.frame(
  name = c('George','John','Paul','Ringo'),
  plays = c('sitar','guitar','bass','drums'), 
  stringsAsFactors=F)
left_join(songs, artists, by='name')
```

### `inner_join` 
(only join things where a and b both have matches)

```{r inner_join}
inner_join(songs, artists, by = "name")
```

### `semi_join` 
(same as inner join, but excludes columns from second table - subset of original table where key is matching)

```{r semi_join}
semi_join(songs, artists, by = "name")
```

### `anti_join` 
(inverse of semi_join, show the rows in the left table that don't have a match in the right table)

```{r anti_join}
anti_join(songs, artists, by = "name")
```

### per capita CO<sub>2</sub> emissions 

You'll join the [population](http://data.okfn.org/data/core/population) dataset to calculate per capita CO<sub>2</sub> emissions.

_**Task**. Summarize the total emissions by country  (not World or EU28) per capita across years from your long format table and return the top 5 emitting countries for **2014**._

_**Task**. Summarize the total emissions by country (not World or EU28) per capita across years from your long format table and return the top 5 emitting countries on **average** across all years._ 

```{r read co2}
library(dplyr)
library(readxl) # install.packages('readxl')

# xls downloaded from http://edgar.jrc.ec.europa.eu/news_docs/CO2_1970-2014_dataset_of_CO2_report_2015.xls
xls = './data/co2_europa.xls'

print(getwd())
co2 = read_excel(xls, skip=12)
names(co2) = c('country', 1970:2014)
  library(dplyr)
  library(tidyr)
co2l = gather(co2,"year","emit", -country)
co2l$year = as.numeric(co2l$year) # Convert "year" column to numeric so that it can be joined to popn
```

```{r popn}
library(readr)
library(tidyr)

popn = read_csv(
  'https://raw.githubusercontent.com/datasets/population/master/data/population.csv')
names(popn) = c('country','code','year','popn')
```


## Task 1 
_**Task**. Summarize the total emissions by country  (not World or EU28) per capita across years from your long format table and return the top 5 emitting countries for **2014**._
```{r Task 1}
library(dplyr)
cpop = left_join(co2l, popn, by=c('country','year'))
cpop %>%
  mutate(pcemit = emit/popn) %>%
  filter(year %in% 2014) %>%
  subset(country != c("World","EU28")) %>%
  arrange(desc(pcemit)) %>%
  head(n=5)
```

```{r}

```

## Task 2
_**Task**. Summarize the total emissions by country (not World or EU28) per capita across years from your long format table and return the top 5 emitting countries on **average** across all years._ 
```{r Task 2}
cpop %>%
  mutate(pcemit = emit/popn) %>%
  group_by(country) %>%
  summarise(avtot = mean(pcemit)) %>%
  arrange(desc(avtot)) %>%
  head(n=5)
  ```
